---
title: "Economic Indicators vs Home-Ownership" 
author: "Kasey Savage"
output: 
  pdf_document: 
  html_document: 
    toc: true 
    toc_depth: 2 
    number_sections: false 
  default date: "2024-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **RESEARCH QUESTION:**

How do changes in macroeconomic indicators(such as unemployment rate, GDP, and Mortgage Rates) influence home-ownership rates across different racial group and regions within the United States of America?

The goal of this analysis is to analyze the causal relationships between broad economic indicators and housing equity in order to drive towards an understanding of if/how macroeconomic forces shape racial and/or regional disparities. By focusing on racial and regional groups, the question highlights structural inequalities often overlooked in aggregate analyses while creating a link between policy and practice.

```{r, echo=TRUE, results="hide", warning=FALSE, message=FALSE}

library(tidyverse)

library(dplyr)

library(httr)

library(readr)

library(stringr)

library(readxl)

library(ggplot2)

library(readr)
```

## **DATA DESCRIPTION:**

**Data comes from a combination of authoritative U.S. and international sources: FRED, US Bureau of Economic Analysis, World Bank, US Office of Management and Budget, U.S. Census Bureau Source, U.S. Department of Housing and Urban Development, Freddie Mac, and a census survey named 'Current Population Survey (Household Survey)'.**

```{r echo=TRUE, results="hide", warning=FALSE, message=FALSE}

GDP <- read_csv("Gross Domestic Product, Billions of Dollars, Quarterly, Seasonally Adjusted Annual Rate.xls - FRED Graph.csv", skip = 10)     # Source: [GDP Data - FRED](https://fred.stlouisfed.org/series/GDP)

Inflation <- read_csv("Inflation, consumer prices for the United States, Percent, Annual, Not Seasonally Adjusted.xls - FRED Graph.csv", skip = 10)     #Source: [Inflation - FRED](https://fred.stlouisfed.org/series/FPCPITOTLZGUSA)

FDGDP <- read_csv("Federal Debt_ Total Public Debt as Percent of Gross Domestic Product, Percent of GDP, Quarterly, Seasonally Adjusted.xls - FRED Graph.csv", skip = 10)     #Source: [Federal Debt - FRED](https://fred.stlouisfed.org/series/GFDEGDQ188S)

UNRATE <- read_csv("UNRATE.csv")     #Source: [Unemployment Rate - FRED](https://fred.stlouisfed.org/series/UNRATE)

MSPUS <- read_csv("MSPUS.csv")     #Source: [Median Sales Price - FRED](https://fred.stlouisfed.org/series/MSPUS)

HOR <- read_csv("RSAHORUSQ156S.csv")     #Source: [Homeownership Rate - FRED](https://fred.stlouisfed.org/series/RSAHORUSQ156S)

BHOR <- read_csv("BOAAAHORUSQ156N.csv")     #Source: [Black Homeownership Rate - FRED](https://fred.stlouisfed.org/series/BOAAAHORUSQ156N) 

AHOR <- read_csv("ANHPIHORUSQ156N.csv")     #Source: [Asian Homeownership Rate - FRED](https://fred.stlouisfed.org/series/ANHPIHORUSQ156N)

OHOR <- read_csv("AORHORUSQ156N.csv")     #Source: [Other Race Homeownership Rate - FRED](https://fred.stlouisfed.org/series/AORHORUSQ156N) 

NHWHOR <- read_csv("NHWAHORUSQ156N.csv")    #Source: [Non-Hispanic White Homeownership - FRED](https://fred.stlouisfed.org/series/NHWAHORUSQ156N

HHOR <- read_csv("HOLHORUSQ156N.csv")    #Source: [Hispanic Homeownership Rate - FRED](https://fred.stlouisfed.org/series/HOLHORUSQ156N)

HORWR <- read_csv("RHORWEQ156N.csv")    #https://fred.stlouisfed.org/series/RHORWEQ156N

HORSR <- read_csv("RHORSOQ156N.csv")    #https://fred.stlouisfed.org/series/RHORSOQ156N

HORNER <- read_csv("RHORNEQ156N.csv")    #https://fred.stlouisfed.org/series/RHORNEQ156N

HORMWR <- read_csv("RHORMWQ156N.csv")    #https://fred.stlouisfed.org/series/RHORMWQ156N

MORTGAGE <- read_csv("MORTGAGE30US-1 - Sheet1 (1).csv")    #Source: [Mortgage Rates - FRED](https://fred.stlouisfed.org/series/MORTGAGE30US)

```

## **DATA CLEANING AND PROCESSING:**

```{r}

GDP <- GDP |> 
  rename(
    `Gross Domestic Product (Billions of Dollars)` = GDP,
    DATE = observation_date
  )

Inflation <- Inflation |> 
  rename(
    `US CP Inflation Percent` = FPCPITOTLZGUSA,
    DATE = observation_date
  )

FDGDP <- FDGDP |> 
  rename(
    `Federal Debt as Percent of GDP` = GFDEGDQ188S,
    DATE = observation_date
  )

UNRATE <- UNRATE |> 
  rename(Unemployment.Rate = UNRATE)

MORTGAGE <- MORTGAGE |> 
  rename(Mortgage.Rates = `30 Year Mortgage Rate`)

MSPUS <- MSPUS |> 
  rename(`Median Sales Price of Houses Sold in USA` = MSPUS)

HOR <- HOR |> 
  rename(`TOTAL HOME OWNERSHIP RATE` = RSAHORUSQ156S)

BHOR <- BHOR |> 
  rename(`HOME OWNERSHIP RATE BLACK` = BOAAAHORUSQ156N)

AHOR <- AHOR |> 
  rename(`HOME OWNERSHIP RATE ASIAN` = ANHPIHORUSQ156N)

HHOR <- HHOR |> 
  rename(`HOME OWNERSHIP RATE HISPANIC` = HOLHORUSQ156N)

NHWHOR <- NHWHOR |> 
  rename(`HOME OWNERSHIP RATE NON-HISPANIC WHITE` = NHWAHORUSQ156N)

OHOR <- OHOR |> 
  rename(`HOME OWNERSHIP RATE OTHER` = AORHORUSQ156N)

HORWR <- HORWR |> 
  rename(`HOME OWNERSHIP RATE WEST` = RHORWEQ156N)

HORSR <- HORSR |> 
  rename(`HOME OWNERSHIP RATE SOUTH` = RHORSOQ156N)

HORNER <- HORNER |> 
  rename(`HOME OWNERSHIP RATE NORTHEAST` = RHORNEQ156N)

HORMWR <- HORMWR |> 
  rename(`HOME OWNERSHIP RATE MIDWEST` = RHORMWQ156N)

```

**Within the merged_data_race data-set I am creating, I am missing 22 values within AHOR(Asian Home-ownership Rate data set). I used a full join here to keep all of the data, just filling the missing values for the Asian data with NAs.**

```{r}

merged_data_race <- GDP |>
  inner_join(Inflation, by = "DATE") |>
  inner_join(FDGDP, by = "DATE") |>
  inner_join(UNRATE, by = "DATE") |>
  inner_join(MSPUS, by = "DATE") |>
  inner_join(HOR, by = "DATE") |>
  inner_join(BHOR, by = "DATE") |>
  full_join(AHOR, by = "DATE") |>
  inner_join(HHOR, by = "DATE") |>
  inner_join(NHWHOR, by = "DATE") |>
  inner_join(OHOR, by = "DATE") |>
  inner_join(MORTGAGE, by = "DATE")

```

```{r}

merged_data_region <- GDP |>
  inner_join(Inflation, by = "DATE") |>
  inner_join(FDGDP, by = "DATE") |>
  inner_join(UNRATE, by = "DATE") |>
  inner_join(MSPUS, by = "DATE") |>
  inner_join(HORWR, by = "DATE") |>
  inner_join(HORSR, by = "DATE") |>
  inner_join(HORNER, by = "DATE") |>
  inner_join(HORMWR, by = "DATE") |>
  inner_join(MORTGAGE, by = "DATE")

```

```{r}

colnames(merged_data_race) <- make.names(colnames(merged_data_race), unique = TRUE)

colnames(merged_data_race)

```

```{r}

colnames(merged_data_region) <- make.names(colnames(merged_data_region), unique = TRUE)

colnames(merged_data_region)

```

```{r}

merged_data_race <- merged_data_race |>
  mutate(across(
    c(Federal.Debt.as.Percent.of.GDP,Gross.Domestic.Product..Billions.of.Dollars., HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE, HOME.OWNERSHIP.RATE.BLACK, HOME.OWNERSHIP.RATE.HISPANIC, HOME.OWNERSHIP.RATE.OTHER, HOME.OWNERSHIP.RATE.ASIAN), 
    ~ (./lag(.) - 1) * 100, 
    .names = "Pct.Change.{.col}"
  )) %>% 
  mutate(Inflation.Growth = US.CP.Inflation.Percent - lag(US.CP.Inflation.Percent))

```

```{r}

merged_data_region <- merged_data_region |>
  mutate(across(
    c(Federal.Debt.as.Percent.of.GDP,Gross.Domestic.Product..Billions.of.Dollars., HOME.OWNERSHIP.RATE.MIDWEST, HOME.OWNERSHIP.RATE.SOUTH, HOME.OWNERSHIP.RATE.NORTHEAST, HOME.OWNERSHIP.RATE.WEST), 
    ~ (./lag(.) - 1) * 100, 
    .names = "Pct.Change.{.col}"
  )) %>% 
  mutate(Inflation.Growth = (US.CP.Inflation.Percent - lag(US.CP.Inflation.Percent)))

```

```{r echo=TRUE, results="hide", warning=FALSE, message=FALSE}

colnames(merged_data_race)

colnames(merged_data_region)

```

```{r}

merged_data_race <- merged_data_race %>%
  mutate(
    DATE = as.Date(DATE),
    across(c(
      Gross.Domestic.Product..Billions.of.Dollars.,
      US.CP.Inflation.Percent,
      Federal.Debt.as.Percent.of.GDP,
      Mortgage.Rates,
      Unemployment.Rate,
      Median.Sales.Price.of.Houses.Sold.in.USA,
      TOTAL.HOME.OWNERSHIP.RATE,
      HOME.OWNERSHIP.RATE.BLACK,
      HOME.OWNERSHIP.RATE.ASIAN,
      HOME.OWNERSHIP.RATE.HISPANIC,
      HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE,
      HOME.OWNERSHIP.RATE.OTHER,
      Inflation.Growth,
      Pct.Change.HOME.OWNERSHIP.RATE.BLACK,
      Pct.Change.HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE,
      Pct.Change.HOME.OWNERSHIP.RATE.HISPANIC,
      Pct.Change.HOME.OWNERSHIP.RATE.ASIAN,
      Pct.Change.HOME.OWNERSHIP.RATE.OTHER,
      Pct.Change.Federal.Debt.as.Percent.of.GDP,
      Pct.Change.Gross.Domestic.Product..Billions.of.Dollars.
    ), as.numeric),
  )

```

```{r}

merged_data_region <- merged_data_region %>%
  mutate(
    DATE = as.Date(DATE),
    across(c(
      Gross.Domestic.Product..Billions.of.Dollars.,
      US.CP.Inflation.Percent,
      Federal.Debt.as.Percent.of.GDP,
      HOME.OWNERSHIP.RATE.WEST,
      HOME.OWNERSHIP.RATE.NORTHEAST,
      HOME.OWNERSHIP.RATE.SOUTH,
      HOME.OWNERSHIP.RATE.MIDWEST,
      Mortgage.Rates,
      Unemployment.Rate,
      Median.Sales.Price.of.Houses.Sold.in.USA,
      Pct.Change.Gross.Domestic.Product..Billions.of.Dollars.,
      Inflation.Growth,
      Pct.Change.HOME.OWNERSHIP.RATE.MIDWEST,
      Pct.Change.HOME.OWNERSHIP.RATE.SOUTH,
      Pct.Change.HOME.OWNERSHIP.RATE.NORTHEAST,
      Pct.Change.HOME.OWNERSHIP.RATE.WEST,
      Pct.Change.Federal.Debt.as.Percent.of.GDP
    ), as.numeric)
  )

```

```{r echo=TRUE, results="hide", warning=FALSE, message=FALSE}

str(merged_data_race)
str(merged_data_region)


```

```{r}

summary(merged_data_race[, c("HOME.OWNERSHIP.RATE.BLACK",        "HOME.OWNERSHIP.RATE.HISPANIC","HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE", "HOME.OWNERSHIP.RATE.ASIAN", "HOME.OWNERSHIP.RATE.OTHER", "TOTAL.HOME.OWNERSHIP.RATE", "Federal.Debt.as.Percent.of.GDP", "Mortgage.Rates", "Pct.Change.HOME.OWNERSHIP.RATE.BLACK",
      "Pct.Change.HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE",
      "Pct.Change.HOME.OWNERSHIP.RATE.HISPANIC",
      "Pct.Change.HOME.OWNERSHIP.RATE.ASIAN",
      "Pct.Change.HOME.OWNERSHIP.RATE.OTHER",
      "Pct.Change.Federal.Debt.as.Percent.of.GDP",
      "Pct.Change.Gross.Domestic.Product..Billions.of.Dollars.")])

```

```{r}

race_data_long <- merged_data_race %>%
  
  select(DATE, 
         HOME.OWNERSHIP.RATE.BLACK, 
         HOME.OWNERSHIP.RATE.HISPANIC, 
         HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE, 
         HOME.OWNERSHIP.RATE.ASIAN, 
         HOME.OWNERSHIP.RATE.OTHER, 
         TOTAL.HOME.OWNERSHIP.RATE) %>%
  
  pivot_longer(
    cols = -DATE, 
    names_to = "Race", 
    values_to = "Homeownership.Rate"
  )

```

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot(merged_data_race, aes(x = DATE)) +
  
  geom_line(aes(y = HOME.OWNERSHIP.RATE.BLACK, color = "Black")) +
  
  geom_line(aes(y = HOME.OWNERSHIP.RATE.HISPANIC, color = "Hispanic")) +
  
  geom_line(aes(y = HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE, color = "White")) +
  
  geom_line(aes(y = HOME.OWNERSHIP.RATE.ASIAN, color = "Asian")) +
  
  geom_line(aes(y = HOME.OWNERSHIP.RATE.OTHER, color = "Other")) +
  
  geom_line(aes(y = TOTAL.HOME.OWNERSHIP.RATE, color = "Total")) +
  
  labs(
    title = "Homeownership Rates For Racial Groups Over Time(YoY)",
    x = "Year",
    y = "Homeownership Rate(%)",
    color = "Race"
  ) +
  
  scale_color_manual(values = c(
    "Black" = "blue",
    "Hispanic" = "red",
    "White" = "yellow",
    "Asian" = "purple",
    "Other" = "orange",
    "Total" = "black"
  )) +
  
  theme_minimal() +
  
  theme(legend.position = "right")

```

**This graph shows the trends in home-ownership rates for different racial groups over time. It plots the home-ownership rates for Black, Hispanic, White, Asian, and Other racial groups, as well as the overall total home-ownership rate, on the y-axis against the years on the x-axis. We can see that the White Race has exceptionally high home-ownership rates, being well above average. Other races and the Asian race have somewhat moderate home-ownership rates, falling just under the average. The Black and Hispanic race fall far below the average. All groups also seem to be following similar trends. There is also much more variability in the minority groups than within the White group.**

```{r echo=TRUE, warning=FALSE, message=FALSE}

ggplot(merged_data_region, aes(x = DATE)) +

geom_line(aes(y = HOME.OWNERSHIP.RATE.MIDWEST, color = "Midwest")) +

geom_line(aes(y = HOME.OWNERSHIP.RATE.WEST, color = "West")) +

geom_line(aes(y = HOME.OWNERSHIP.RATE.NORTHEAST, color = "Northeast")) +

geom_line(aes(y = HOME.OWNERSHIP.RATE.SOUTH, color = "South")) +

labs( title = "Homeownership Rates For USA Regions Over Time(YoY)", x = "Year", y = "Homeownership Rate(%)", color = "Region" ) +

scale_color_manual(values = c( "Midwest" = "blue", "West" = "red", "South" = "yellow", "Northeast" = "purple")) +

theme_minimal() +

theme(legend.position = "right")

```

**This graph shows the trends in home-ownership rates for different USA regional groups over time. It plots the home-ownership rates for the Midwest, West, Northeast, and South on the y-axis against the years on the x-axis. We can see that the Midwest region and the South region have fairly hig home-ownership rates in comparison to the Northeast and West regions. All regions also seem to be following similar trends, though, there is more variability in the Northeast and West regions.**

## **STATISTICAL ANALYSIS**

### **Race Analysis:**

```{r, echo=TRUE, warning=FALSE, message=FALSE}

library(fixest)


model1 <- feols(c(HOME.OWNERSHIP.RATE.BLACK, 
         HOME.OWNERSHIP.RATE.HISPANIC, 
         HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE, 
         HOME.OWNERSHIP.RATE.ASIAN, 
         HOME.OWNERSHIP.RATE.OTHER) ~
        Mortgage.Rates + Inflation.Growth + Unemployment.Rate +Median.Sales.Price.of.Houses.Sold.in.USA + Federal.Debt.as.Percent.of.GDP, data = merged_data_race)

summary(model1)

```

***\*I will be using a significance level of .05 (α = 0.05) for this statistical analysis.***\*

**Black Race Analysis:** *For Black home-ownership rates, the intercept indicates a base rate of 46.18%* when all other factors are at zero(statistically significant)*. Mortgage rates and inflation growth do not show statistically significant relationships with Black home-ownership. The unemployment rate, however, is positively associated, with an increase of 0.58% per 1% increase The median sales price of houses sold in the U.S. shows a positive relationship, with home-ownership increasing by 0.000042 percentage points per dollar. Conversely, federal debt as a percentage of GDP is significantly associated with a decrease of 0.19% in home-ownership rate per 1% increase.*

**Hispanic Race Analysis:** *For Hispanic home-ownership rates, the intercept indicates a base rate of 45.34%* when all other factors are at zero(statistically significant)*. Mortgage rates and inflation growth do not show statistically significant relationships with Hispanic home-ownership. The unemployment rate has a positive association, with an increase of 0.53% in home-ownership per 1% increase in unemployment. The median sales price of houses sold in the U.S. is positively associated, increasing home-ownership by 0.000055% per dollar. Federal debt as a percentage of GDP is significantly associated with a decrease of 0.16% in home-ownership rate for a 1% increase.*

**Non-Hispanic White Race Analysis:** *For non-Hispanic White home-ownership rates, the intercept indicates a base rate of 74.64%* when all other factors are at zero(statistically significant)*. Mortgage rates have a marginally significant negative relationship, with a decrease of 0.28% for each 1% increase in mortgage rates. Inflation growth does not show a statistically significant relationship. The unemployment rate has a strong positive association, with a 0.53% increase in home-ownership for each 1% increase in unemployment. The median sales price of houses sold in the U.S. is positively associated, increasing home-ownership by 0.000034% per dollar. Federal debt as a percentage of GDP is also significantly associated with a decrease in home-ownership, with a reduction of 0.13% for a 1% increase in Federal debt as a percentage of GDP.*

**Asian Race Analysis:** *For Asian home-ownership rates, the intercept suggests a base rate of 46.67%* when all other factors are at zero(statistically significant)*. All other results were not statistically significant at the 5% level.*

**Other Race Analysis**: *For home-ownership rates among individuals classified as 'Other,' the intercept indicates a base home-ownership rate of 55.59%* when all other factors are at zero(statistically significant)*. Mortgage rates have a statistically significant negative impact, with a decrease of 0.68% for each 1% increase in mortgage rates. Unemployment rate, median sales price of houses, and federal debt as a percentage of GDP all significantly affect home-ownership rates, with increases in unemployment and home prices leading to higher rates of home-ownership, while higher federal debt as a percentage of GDP decreases home-ownership. Inflation growth, however, does not show a significant effect on home-ownership rates. Overall, mortgage rates, unemployment, housing prices, and federal debt are important predictors of home-ownership rates for this group.*

### **Pairs Plot**

**The pair plots will illustrate the relationships between macroeconomic indicators and home-ownership rates across different racial groups, highlighting potential correlations and patterns in the data.**

```{r, echo=TRUE, results="hide", warning=FALSE, message=FALSE}

library(GGally)

```

```{r echo=TRUE, warning=FALSE, message=FALSE}

selected_data1 <- merged_data_race[, c("Mortgage.Rates", "Inflation.Growth", "Unemployment.Rate", "Median.Sales.Price.of.Houses.Sold.in.USA", "Federal.Debt.as.Percent.of.GDP", "HOME.OWNERSHIP.RATE.BLACK", "HOME.OWNERSHIP.RATE.HISPANIC", "HOME.OWNERSHIP.RATE.NON.HISPANIC.WHITE","HOME.OWNERSHIP.RATE.ASIAN","HOME.OWNERSHIP.RATE.OTHER")]

ggpairs(selected_data1, lower = list(continuous = wrap("points", color = "red")), 
        upper = list(continuous = wrap("cor", size = 3)))+ 
  
  theme( strip.text = element_text(size = 6), 
         axis.text = element_text(size = 6),
         axis.title = element_text(size = 5), 
         legend.text = element_text(size = 6))
        
```

### Key Observations:

-   The correlation between Home-Ownership Rate and Median Sales Price of Houses varied between racial groups. For Asian, Hispanic, and Other racial groups, there was a strong positive relation, whle for the Non-Hispanic White group there was a moderate positive correlation. The Black racial group, however, had a negative and moderate correlation. This suggests that rising housing prices correlate with higher home-ownership rates among all groups except for the Black race group. Ex: Correlation for "HOME.OWNERSHIP.RATE.BLACK" and "Median.Sales.Price" is -0.238\*\*\*.

-   Federal Debt as Percent of GDP had a positive correlation with home-ownership for most racial groups. There was a moderate negative correlation for the Black group, and a weak negative correlation for the Non-Hispanic White group. This may indicate that rising national debt correlates with more favorable ownership conditions for certain groups.

-   For the races Hispanic, White, and Other, Mortgage Rates had a negative correlation. For Asians and Blacks, there was a positive correlation.

-   There were strong positive correlations that exist between different racial groups' home-ownership rates, indicating shared trends over time. An example is "HOME.OWNERSHIP.RATE.BLACK" and "HOME.OWNERSHIP.RATE.ASIAN": 0.942\*\*\*.

### Regional Analysis:

```{r, echo=TRUE, results="hide", warning=FALSE, message=FALSE}

colnames(merged_data_region)

```

```{r, echo=TRUE, warning=FALSE, message=FALSE}

model2 <- feols(c(HOME.OWNERSHIP.RATE.WEST, 
         HOME.OWNERSHIP.RATE.SOUTH, 
         HOME.OWNERSHIP.RATE.MIDWEST, 
         HOME.OWNERSHIP.RATE.NORTHEAST) ~
        Mortgage.Rates + Inflation.Growth + Unemployment.Rate +Median.Sales.Price.of.Houses.Sold.in.USA + Federal.Debt.as.Percent.of.GDP, data = merged_data_region)

summary(model2)

```

**West Region Analysis:** *The analysis reveals that the intercept, representing the base home-ownership rate in the West when all other factors are at zero, is estimated at 66.94%. Mortgage rates, federal debt as a percentage of GDP, and median sales prices of houses are significantly associated with home-ownership rates in the West. Higher mortgage rates and federal debt correspond to lower home-ownership, while higher median house prices are linked to increased home-ownership, potentially reflecting economic prosperity in the region. Unemployment rates also show a positive and significant relationship with home-ownership, a result that may suggest demographic or policy-driven dynamics. Inflation growth does not significantly affect home-ownership rates in this analysis*

**South Region Analysis** *The analysis indicates that the intercept, representing the base home-ownership rate in the South when all other factors are at zero, is estimated at 71.55%. Mortgage rates and federal debt as a percentage of GDP have significant negative associations with home-ownership, while higher median house prices are significantly linked to increased home-ownership rates, potentially reflecting broader economic dynamics. Unemployment rates also have a positive and significant association with home-ownership, suggesting regional factors or policies that may influence housing outcomes. Inflation growth does not exhibit a significant relationship with home-ownership rates in the South.*

**Midwest Region Analysis** *The intercept suggests a baseline home-ownership rate of 75.99% in the Midwest when all other variables are held constant. Mortgage rates have a significant negative effect on home-ownership, with a decrease of 0.47% for each percentage-point increase in rates. Federal debt as a percentage of GDP also negatively impacts home-ownership, reducing it by 0.17% per 1% increase. Median sales price of houses has a small but statistically significant positive relationship, increasing home-ownership by 0.000036% for every dollar rise. Unemployment rates show a positive effect, while inflation growth does not significantly influence home-ownership in this region.*

**Northeast Region Analysis** *In the Northeast, the intercept suggests a baseline home-ownership rate of 63.14%. Mortgage rates are associated with a significant decrease of 0.27% per 1% increase, while federal debt as a percentage of GDP is linked to a decrease of 0.11%. Median sales price of houses sold in the U.S. shows a positive relationship, with home-ownership increasing by 0.00003 percentage points per dollar. Unemployment rate is positively associated with home-ownership, increasing by 0.41% per 1% increase, whereas inflation growth does not have a significant effect.*

### **Pairs Plot**

**The pair plots will illustrate the relationships between macroeconomic indicators and home-ownership rates across different regional groups, highlighting potential correlations and patterns in the data.**

```{r, echo=TRUE, warning=FALSE, message=FALSE}

selected_data <- merged_data_region[, c("Mortgage.Rates", "Inflation.Growth", "Unemployment.Rate", "Median.Sales.Price.of.Houses.Sold.in.USA", "Federal.Debt.as.Percent.of.GDP", "HOME.OWNERSHIP.RATE.MIDWEST", "HOME.OWNERSHIP.RATE.WEST", "HOME.OWNERSHIP.RATE.SOUTH", "HOME.OWNERSHIP.RATE.NORTHEAST")]

ggpairs(selected_data, lower = list(continuous = wrap("points", color = "blue")), 
        upper = list(continuous = wrap("cor", size = 3))) + 
  theme( strip.text = element_text(size = 6), 
         axis.text = element_text(size = 6),
         axis.title = element_text(size = 5), 
         legend.text = element_text(size = 6))
        
        
```

### Key Observations:

-   The correlations between Median Sales Price of Houses and Home-Ownership Rates across regions showed positive correlations, moderately strong in the West and Northeast. Ex: Correlation between Median Sales Price vs. HOME.OWNERSHIP.RATE.NORTHEAST: 0.490\*\*. This suggests that rising housing prices help home-ownership more in these regions.

-   The correlations between Unemployment Rate and Home-Ownership Rate across all regions showed negative but weaker correlations, meaning lower unemployment may still impact ownership but less directly.

-   Mortgage Rates generally had weaker correlations with home-ownership across all regions, but moderatley negative.

-   There were strong positive correlations that exist between different regional groups' home-ownership rates, indicating shared trends over time. This is seen especially between the South and Midwest (e.g., 0.857\*\*\*), indicating that regional home-ownership rates often follow similar patterns.

## **CHALLENGES:**

One big challenge I ran into was dealing with inconsistent scaling across the data-sets. Some data-sets had weekly data while others used quarterly or annual data. This made it tough to compare things directly. Another issue was handling missing data, which is pretty common in economic data-sets due to reporting lags, or because certain years and/or demographics weren't included. The data on the Asian race group had 22 missing values, and this made it hard to perform statistical test on this group, and accurately compare them to the other groups. I also wanted to create interactive terms, however, finding a way to do this that made sense and was easy to interpret was quite difficult, so I strayed away from it. Lastly, some of the data-sets came in different formats and units, like percentages versus absolute values. This took time to get aligned.

## **DISCUSSION:**

In this analysis, I explored how changes in macroeconomic indicators—such as GDP, unemployment rate, inflation, and mortgage rates—impact home-ownership rates across different racial groups and regions in the United States. The goal was to uncover whether broad economic forces exacerbate racial and regional disparities in housing equity. This is what I found.

### Race Discussion:

Home-ownership rates across racial groups reveals notable disparities, with the Black and Hispanic communities typically experiencing lower home-ownership rates compared to Non-Hispanic Whites, Asians, and Other races. Mortgage rates, inflation growth, unemployment, and median sales prices are significant factors affecting home-ownership, yet their impacts are not uniform across racial groups. The Black and Hispanic populations appear to be more vulnerable to shifts in economic conditions, underscoring structural barriers that hinder equitable access to home-ownership. The statistical analysis I performed shows that while economic indicators influence all racial groups, the degree to which these factors affect home-ownership rates varies, highlighting the persistent racial inequities in housing. Another important discovery from this analysis was how unemployment rates are sometimes associated with higher home-ownership rates, especially for Black and Hispanic groups. While this seems counter-intuitive, it allows us to understand that other factors play a huge role in home-ownership rates outside of the economic state of the country. While the data does not include direct information on government policies such as foreclosure moratoriums(ex. FHA paused foreclosures from March 2020-July 2021), you can hypothesize that these types of policies could have played a role in mitigating some negative impacts during times of high unemployment, especially for minorities.\*

### Region Discussion:

When examining regional differences, id like to begin by saying that, like the race data, while the majority of macroeconomic factors used within this analysis does show correlation to housing prices for these groups, there seem to be plenty of other factors that drive these home-ownership rates as well. We are able to see that the variation in home-ownership rates between regions reflects not only the economic differences but also differences that were not covered within this analysis, such as the cultural and historical factors that shape housing markets. However, we are able to see some important correlations from the data. The data suggests that the South and Midwest regions have relatively higher home-ownership rates compared to the Northeast and West regions. Economic indicators, such as mortgage rates and unemployment, influence these regional differences. For example, the Midwest region has a more stable home-ownership environment, likely due to a lower cost of living and more affordable housing. In contrast, the West region, which often faces higher housing costs and a more volatile job market, experiences greater fluctuations in home-ownership rates. We are also able to see that rising housing prices hurt home-ownership more in the South and Midwest than in the Northeast and West.
